\titletext{}

### 2021-06-21 manuscript/regional_befr_manuscript.tex
Finished revisions

\usepackage{xcolor}
\newcommand{\TODO}[1]{\textcolor{red}{\textbf{#1}}}  % \TODO{NOTE TO SELF WRITTEN IN RED}

@@

\begin{itemize}
	\item{Positive biodiversity-ecosystem function relationships (BEFRs) have been widely documented, but it is unclear if BEFRs should be expected in disturbance-driven systems. Disturbance may limit competition and niche differentiation, which are frequently posited to underlie BEFRs. We provide the first exploration of the relationship between tree species diversity and biomass, one measure of ecosystem function, across southern African woodlands and savannas, an ecological system rife with disturbance from fire, herbivores and humans.}
	\item{We used >1000 vegetation plots distributed across 10 southern African countries, and structural equation modelling, to determine the relationship between tree species diversity and aboveground woody biomass, accounting for interacting effects of resource availability, disturbance by fire, tree stem density at all and vegetation type.}
	\item{We found positive effects of tree species diversity on aboveground biomass, operating via increased structural diversity. The observed BEFR was highly dependent on organismal density, with a minimum threshold of c. 180 mature stems ha\textsuperscript{-1}. We found that water availability mainly affects biomass indirectly, via increasing species diversity.}
	\item{The study underlines the close association between tree diversity, ecosystem structure, environment and function in highly disturbed savannas and woodlands. We suggest that tree diversity is an under-appreciated determinant of wooded ecosystem structure and function.}
\end{itemize}

@@

A number of studies in miombo woodlands, the dominant savanna type in southern Africa \citep{Campbell1996}, have found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. These studies however, lacked the spatial extent required to account for the interacting effects of variation in abiotic environment, disturbance regime, and biogeography which are expected to affect the BEFR. Studies of the BEFR often find that within a small spatial extent (<2500 km\textsuperscript{2}), biodiversity shows a strong effect on ecosystem function, but at broader extents (>100,000s km\textsuperscript{2}) biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Gonzalez2020}. In West Africa, \citet{Mensah2020} found that woodlands and forests showed a positive effect of tree species richness on above-ground carbon, while sparse savannas did not, implying that tree stem density and the presence of dominant large trees may affect the strength of the observed BEFR. In the Brazilian Cerrado savanna, \citet{Loiola2015} found that disturbance by fire reduced tree productivity via its effect on functional trait values. In contrast however, \citet{Carvalho2014} also working in the Brazilian Cerrado, found that while disturbance by fire did reduce soil fertility, which is often closely related to productivity in savannas, functional trait diversity was unaffected by fire frequency. Other studies focussing on the herbaceous diversity in disturbance-prone grasslands in North America and Europe have shown that disturbance by fire, mowing and herbivory reduces herbaceous productivity \citep{Grace2007}, allowing weak competitors to co-exist where otherwise they would be excluded \citep{Mason2011}, thus weakening the observable BEFR. Thus, there is a case that in highly disturbed systems, a relationship between diversity and ecosystem function may not exist at all. While these studies together offer some glimpse into how the BEFR may operate in disturbance-prone systems, none provide a full and comprehensive assessment of the interacting effects of diversity, abiotic environment, disturbance and ecosystem function, particularly for wooded ecosystems. Additionally, due to differences in community assembly, evolutionary history and contemporary drivers between the neotropics and the African tropics, inferences from one continent cannot necessarily be applied to the other \citep{Dexter2015}.

% In this study :::
In this study, we make the first known estimation of the Biodiversity-Ecosystem Function Relationship (BEFR) in disturbance-driven wooded ecosystems (savannas and woodlands), using southern Africa as our study region. We aim to understand the synergistic effects of environmental and biotic drivers of variation in the BEFR, with a view to creating a general model of the BEFR in disturbance-prone wooded ecosystems, which is currently lacking in the BEFR literature. We posit three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect of tree species diversity on woody biomass will increase with organismal density (number of stems ha\textsuperscript{-1}), with competitive interactions becoming more pronounced as stem proximity increases. Relatedly, we expect that an increase in disturbance by fire will decrease organismal density and therefore competition, weakening the effect of tree species diversity on woody biomass. Finally, we expect that (3) tree species diversity will increase tree structural diversity due to inter-specific variation in functional form (i.e. physiognomic diversity), providing an indirect path by which tree diversity increases woody biomass.

@@

% Number of stems
All stems >10 cm DBH were measured within each plot resulting in a total of \num[group-separator={,}]{\ntrees{}} stems with measurements. A tree may be comprised of multiple stems and so tree-level richness estimates, rather than stem-level estimates, were used to prevent bias from species which readily coppice. For each tree, we recorded species, DBH and tree height to the top of the highest branch material. Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks. When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m. The height of this measurement was recorded and used to estimate the D\textsubscript{e} at 1.3 m using a cubic polynomial regression, with parameters estimated using a test dataset from Ryan C., (unpublished), see \citet{Godlee2020}.

% AGB Calculation
AGB for each plot (t ha\textsuperscript{-1}) was calculated using \autoref{chave_agb}, taken from \citet{Chave2014}:

\begin{equation}
	AGB = 0.0673 \times (\rho D_{e}^{2} H)^{0.976}
	\label{chave_agb}
\end{equation}

where $\rho$ is the species mean wood density (g cm\textsuperscript{-3}), $D_{e}$ is the estimated DBH (cm) at 1.3 m, and $H$ is the tree height (m). Wood density estimates were taken from the global wood density database for each species where possible \citep{Chave2009, Zanne2009}. Wood density for species without species level estimates was estimated from the means of their respective genera. For stems where tree height was unknown, the plots' climatic parameters, estimated from plot location, were used to estimate tree height, according to \citet{Chave2014}.

@@

We assessed the role of tree species diversity and tree structural diversity in determining AGB via a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity. Structural diversity can also directly influence AGB in this model, separate to the effect of of species diversity. To account for variation in stem density, which may covary with species diversity, we included it as an observed variable in our model. To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation type separately. We compared unstandardised path coefficients among the models for different vegetation types to understand the effect that vegetation type has on the relationship between tree species diversity, structural diversity, stem density and AGB. Path coefficients show the effect of a given path with other paths held constant. Models were estimated using the `MLM' estimator, because it is robust to multivariate non-normality \citep{Shapiro1983}. Model fit was evaluated using the robust Comparative Fit Index (CFI), the robust Tucker Lewis Index (TLI), the Root Mean Squared Error of Approximation (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB. We critically assessed model fit in each case, taking into consideration the recommendations of \citet{Hu1999} who define threshold values of acceptability for these model fit indices: CFI >0.85, TLI >0.85, RMSEA <0.15, alongside our judgement of the model estimates.

@@

Preliminary models that included herbivore biomass \citep{Hempson2017} did not converge. This is possibly due to the spatially coarse nature of the available data, or due to collinearity with other variables, notably MAP and fire frequency. We therefore did not include herbivory in our final model. We incorporated environmental covariates into our model to understand the relative effects of water availability, soil fertility and disturbance on AGB both directly and indirectly via species diversity and stem density. We compared standardised path coefficients among paths in the model to understand the relative contribution of each path to explain variance in AGB. Vegetation type specific models could not be reliably fitted for this more complex model specification with environmental covariates, due to sample size issues and because some vegetation types were narrow in their climate space, leading to a lack of environmental variation, particularly in the water availability latent variable.

@@

In an SEM describing the effect of tree species diversity on AGB via the mediating effects of tree structural diversity and stem density (\autoref{struc_mod}), species diversity showed no direct effect on AGB (\strucbetadb{}), but did have an indirect positive effect via structural diversity (\strucbetadib{}) (\autoref{struc_mod}). Model fit was good with high factor loadings for all observed variables. All other path coefficients were significant (p <0.01) (\autoref{struc_model_fit_clust_stats}). The R\textsuperscript{2} of AGB was \strucrsq{}. The strongest direct effect on AGB was from stem density (\strucbetaib{}).

@@

We found evidence that tree species diversity led to an increase in AGB indirectly via tree structural diversity, and we therefore find support for our second hypothesis H\textsubscript{2}. A higher tree species diversity allows for a greater structural diversity of trees, i.e. greater variation in DBH and height. This may act as a mechanism for niche complementarity, with a canopy of diversely-sized trees able to take advantage of a greater proportion of the available light. Although we did not measure them here, we would also expect that tree species diversity allows for a greater range of tree functional forms \citep{Pretzsch2014}, i.e. wider variation in canopy shape and overall growth form; broad flat crowns vs. narrow deep crowns, for example. In forests, where the tree canopy is effectively closed, as the stand matures a more diverse canopy emerges via competition and tree mortality events which open canopy gaps \citep{Muscolo2014}. Indeed, our finding that the strength of the effect of tree diversity on AGB increases with stem density supports this mechanism (\autoref{sem_struc_stems_ha}). At low stem densities, competition between mature trees may not occur, meaning that the niche complementarity effect provided by an increase in tree species richness may not be present, accounting for the small effect of tree species diversity on AGB below c. 180 trees ha\textsuperscript{-1}. In frequently disturbed woodlands such as those studied here, a woodland canopy similar to that of a forest is frequently not reached. Instead, a simple open canopy is maintained that can be made more complex and productive via an increase in species diversity.

@@

As time since disturbance increases in any wooded ecosystem, it is expected that both species diversity and above-ground biomass will increase until an equilibrium is reached or another disturbance event occurs \citep{Huston2014}. Unless this effect is accounted for, there is the potential for mistaken causality as both diversity and biomass will correlate. In our model, time since disturbance is accounted for within each plot via the stem density term. Disturbance reduces stem density of large stems (>10 cm DBH), which is expected to increase until the effects of competition preclude further increase \citep{Johnson2012}. Furthermore, our rarefied measure of species diversity accounts for variation in sampling effort and is therefore independent of stem density. Tree species richness should also increase with time since disturbance as with increased stem density the likelihood of including a new species also increases. Outside of the stem density effect, there are multiple causes for variation in tree species diversity in this study. Vegetation types and localities differ in their available species pool, for example. Variation in abiotic environmental factors will also affect species accumulation.

@@

A negative total effect of soil fertility on AGB is in contrast to other studies in the region and general ecological theory, which predicts a positive effect of soil nutrients on biomass \citep{Scarascia2000}. The negative total effect of soil fertility on AGB was driven mostly by an indirect negative effect via stem density. The direct effect on AGB however, remained positive and marginally significant, as expected. Model estimates of the effect of soil on AGB were poorly constrained compared with other latent variables. This wide standard error on the model predictions is possibly due to the coarseness and nature of the soil data we used. SoilGrids provides modelled data at 250 m resolution, while soil structure and nutrient content varies at much finer scales in southern African woodlands \citep{Muledi2017, Bucini2007}. It is therefore not surprising that this model path is poorly constrained. \citet{Lehmann2014} found similarly weak and poorly constrained relationships for soil in a Structural Equation Model including precipitation, temperature, soil, and fire to predict tree basal area in southern African woodlands. Plot-specific soil data are time-consuming to collect and difficult to compare across studies when different protocols are used. Our study points to the need for further effort in this regard, which may reveal interesting findings about the complex interactions between soil, disturbance and tree diversity in southern African woodlands.

@@

In this study we found that even in highly disturbed southern African woodlands, there exists a generalisable positive association between tree species diversity and ecosystem function, quantified as above-ground woody biomass (AGB). Our findings contribute to our understanding of a universal biodiversity-ecosystem function relationship, one which is moderated in a predictable manner by environmental covariates and their interaction with biodiversity and ecosystem structure. We found that the multiple vegetation types which comprise southern African woodlands exhibit similarities in the relationship between species diversity and woody biomass, suggesting that similar processes operate across the region to determine ecosystem function. We advocate for explicit inclusion of environmental covariates in regional scale models of biodiversity and ecosystem function. We assert that this is necessary to develop our understanding of the biodiversity-ecosystem function relationship in real-world ecosystems, to progress from experimental mesocosms. We found that much of the effect of species diversity on biomass exists as an indirect effect by increasing the structural diversity of trees, exemplifying a key mechanism by which tree species diversity determines ecosystem function in savannas, woodlands and forests, i.e. all wooded ecosystems. The presence of a stem density threshold above which the effect of tree species diversity on AGB increases clearly implies the presence of niche complementarity effects in southern African woodlands, an aspect which has often been overlooked in previous studies despite its intuitive logic as a determinant of niche complementarity effects in wooded ecosystems. Our study shows that biodiversity change through extensive human actions in this region will have the greatest negative impact on ecosystem function in areas of high stem density, and low species diversity, which are those areas predominantly targeted for tree felling. This raises concerns about the robustness of these ecosystems to further resource extraction and biodiversity loss. Finally, our results provide further evidence of the complex interaction of factors governing biomass and therefore carbon dynamics in disturbance-driven wooded ecosystems, which currently represent the greatest uncertainty in the global terrestrial carbon sink.

>>>>

\usepackage{xcolor}
\newcommand{\todo}[1]{\textcolor{red}{\textbf{#1}}}  % \todo{NOTE TO SELF WRITTEN IN RED}

@@

\begin{itemize}
	\item{Positive biodiversity-ecosystem function relationships (BEFRs) have been widely documented, but it is unclear if BEFRs should be expected in disturbance-driven systems. Disturbance may limit competition and niche differentiation, which are frequently posited to underlie BEFRs. We provide the first exploration of the relationship between tree species diversity and biomass, one measure of ecosystem function, across southern African woodlands and savannas, an ecological system rife with disturbance from fire, herbivores and humans.}
	\item{We used >1000 vegetation plots distributed across 10 southern African countries, and structural equation modelling, to determine the relationship between tree species diversity and aboveground woody biomass, accounting for interacting effects of resource availability, disturbance by fire, tree stem density \rold{at all }and vegetation type.}
	\item{We found positive effects of tree species diversity on aboveground biomass, operating via increased structural diversity. The observed BEFR was highly dependent on organismal density, with a minimum threshold of c. 180 mature stems ha\textsuperscript{-1}. We found that water availability mainly affects biomass indirectly, via increasing species diversity.}
	\item{The study underlines the close association between tree diversity, ecosystem structure, environment and function in highly disturbed savannas and woodlands. We suggest that tree diversity is an under-appreciated determinant of wooded ecosystem structure and function.}
\end{itemize}

@@

A number of studies in miombo woodlands, the dominant savanna type in southern Africa \citep{Campbell1996}, have found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. These studies however, lacked the spatial extent required to account for the interacting effects of variation in abiotic environment, disturbance regime, and biogeography which are expected to affect the BEFR. Studies of the BEFR often find that within a small spatial extent (<2500 km\textsuperscript{2}), biodiversity shows a strong effect on ecosystem function, but at broader extents (>100,000s km\textsuperscript{2}) biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Gonzalez2020}. In West Africa, \citet{Mensah2020} found that woodlands and forests showed a positive effect of tree species richness on above-ground carbon, while sparse savannas did not, implying that tree stem density and the presence of dominant large trees may affect the strength of the observed BEFR. In the Brazilian Cerrado savanna, \citet{Loiola2015} found that disturbance by fire reduced tree productivity via its effect on functional trait values. \rold{In contrast however, }\citet{Carvalho2014}\rold{ also working in the Brazilian Cerrado, found that while disturbance by fire did reduce soil fertility, which is often closely related to productivity in savannas, functional trait diversity was unaffected by fire frequency.}\rnew{Furthermore, }\citet{Carvalho2014}\rnew{ found that disturbance by fire in the Brazilian Cerrado reduced soil fertility, causing an indirect effect of fire frequency on functional trait diversity via soil fertility.} Other studies focussing on the herbaceous diversity in disturbance-prone grasslands in North America and Europe have shown that disturbance by fire, mowing and herbivory reduces herbaceous productivity \citep{Grace2007}, allowing weak competitors to co-exist where otherwise they would be excluded \citep{Mason2011}, thus weakening the observable BEFR. Thus, there is a case that in highly disturbed systems, a relationship between diversity and ecosystem function may not exist at all. While these studies together offer some glimpse into how the BEFR may operate in disturbance-prone systems, none provide a full and comprehensive assessment of the interacting effects of diversity, abiotic environment, disturbance and ecosystem function, particularly for wooded ecosystems. Additionally, due to differences in community assembly, evolutionary history and contemporary drivers between the neotropics and the African tropics, inferences from one continent cannot necessarily be applied to the other \citep{Dexter2015}.

% In this study :::
In this study, we make the first known estimation of the Biodiversity-Ecosystem Function Relationship (BEFR) in disturbance-driven wooded ecosystems (savannas and woodlands), using southern Africa as our study region. We aim to understand the synergistic effects of environmental and biotic drivers of variation in the BEFR, with a view to creating a general model of the BEFR in disturbance-prone wooded ecosystems, which is currently lacking in the BEFR literature. We posit three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect of tree species diversity on woody biomass will increase with organismal density (number of stems ha\textsuperscript{-1}), with competitive interactions becoming more pronounced as stem proximity increases. Relatedly, we expect that an increase in disturbance by fire will decrease organismal density and therefore competition, weakening the effect of tree species diversity on woody biomass. Finally, we expect that (3) tree species diversity will increase tree structural diversity due to inter-specific variation in \rchange{functional form}{size strategy} (i.e. physiognomic diversity), providing an indirect path by which tree diversity increases woody biomass.

@@

% Number of stems
All stems >10 cm DBH were measured within each plot resulting in a total of \num[group-separator={,}]{\ntrees{}} stems with measurements. A tree may be comprised of multiple stems and so tree-level richness estimates, rather than stem-level estimates, were used to prevent bias from species which readily coppice. For each tree, we recorded species, \rold{DBH and }tree height to the top of the highest branch material, \rnew{and the DBH of each stem >10 cm DBH}. Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks. When \rnew{stem }DBH could not be reliably measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m. The height of this measurement was used to estimate diameter at 1.3 m using a cubic polynomial regression to account for stem taper, with parameters estimated using a test dataset from Ryan C., (unpublished), see \citet{Godlee2020}.

% AGB Calculation
AGB for each plot (t ha\textsuperscript{-1}) was calculated \rnew{as the sum of the AGB of each stem >5 cm DBH ($AGB_{i}$), divided by the plot area,} using \autoref{chave_agb} taken from \citet{Chave2014}:

\begin{equation}
	AGB_{i} = 0.0673 \times (\rho D^{2} H)^{0.976} \times 1000
	\label{chave_agb}
\end{equation}

where $\rho$ is the species mean wood density (g cm\textsuperscript{-3}), \rchange{$D_{e}$ is the estimated DBH (cm) at 1.3 m}{$D$ is the stem diameter (cm) measured or estimated at 1.3 m}, and $H$ is the tree height (m). Wood density estimates were taken from the global wood density database for each species where possible \citep{Chave2009, Zanne2009}. Wood density for species without species level estimates was estimated from the means of their respective genera. For stems where tree height was unknown, the plots' climatic parameters, estimated from plot location, were used to estimate tree height, according to \citet{Chave2014}.

@@

\rchange{We assessed the role of tree species diversity and tree structural diversity in determining AGB via a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity. Structural diversity can also directly influence AGB in this model, separate to the effect \rchange{of of}{of} species diversity. To account for variation in stem density, which may covary with species diversity, we included it as an observed variable in our model.}{First, we used a simple mediation model which excluded the environmental covariates, to assess the role of tree species diversity and tree structural diversity in determining AGB. This model allowed direct effects of species diversity, structural diversity, and stem density on AGB, and also the indirect effect of species diversity on AGB via structural diversity.} To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation type separately. We compared unstandardised path coefficients among the models for different vegetation types to understand the effect that vegetation type has on the relationship between tree species diversity, structural diversity, stem density and AGB. Path coefficients show the effect of a given path with other paths held constant. Models were estimated using the `MLM' estimator, because it is robust to multivariate non-normality \citep{Shapiro1983}. Model fit was evaluated using the robust Comparative Fit Index (CFI), the robust Tucker Lewis Index (TLI), the Root Mean Squared Error of Approximation (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB. We critically assessed model fit in each case, taking into consideration the recommendations of \citet{Hu1999} who define threshold values of acceptability for these model fit indices: CFI >0.85, TLI >0.85, RMSEA <0.15, alongside our judgement of the model estimates.

@@

\rold{Preliminary models that included herbivore biomass }\citep{Hempson2017} \rold{did not converge. This is possibly due to the spatially coarse nature of the available data, or due to collinearity with other variables, notably MAP and fire frequency. We therefore did not include herbivory in our final model. We incorporated environmental covariates into our model to understand the relative effects of water availability, soil fertility and disturbance on AGB both directly and indirectly via species diversity and stem density. We compared standardised path coefficients among paths in the model to understand the relative contribution of each path to explain variance in AGB. Vegetation type specific models could not be reliably fitted for this more complex model specification with environmental covariates, due to sample size issues and because some vegetation types were narrow in their climate space, leading to a lack of environmental variation, particularly in the water availability latent variable.}\rnew{Second, we fitted the full model with environmental covariates, to understand the relative effects of water availability, soil fertility and disturbance on AGB, both directly and indirectly via species diversity and stem density. We compared standardised path coefficients among paths in the model to understand the relative contribution of each path to explain variance in AGB. Due to sample size issues, and because some vegetation types were narrow in their climate space, particularly in the water availability latent variable, we could not fit the model including environmental covariates separately for each vegetation type, as we encountered issues with model convergence. Preliminary models that included herbivore biomass }\citep{Hempson2017} \rnew{did not converge. This is possibly due to the spatially coarse nature of the available data, or to collinearity with other variables, notably MAP and fire frequency. We therefore did not include herbivory in our final model.}

@@

\rchange{In an SEM describing the effect of tree species diversity on AGB via the mediating effects of tree structural diversity and stem density}{In the reduced SEM, which included stem density and the mediating effect of species diversity on AGB via structural diversity} (\autoref{struc_mod}), species diversity showed no direct effect on AGB (\strucbetadb{}), but did have an indirect positive effect via structural diversity (\strucbetadib{}) (\autoref{struc_mod}). Model fit was good with high factor loadings for all observed variables. All other path coefficients were significant (p <0.01) (\autoref{struc_model_fit_clust_stats}). The R\textsuperscript{2} of AGB was \strucrsq{}. The strongest direct effect on AGB was from stem density (\strucbetaib{}).

@@

We found evidence that tree species diversity led to an increase in AGB indirectly via tree structural diversity, and we therefore find support for our second hypothesis H\textsubscript{2}. A higher tree species diversity allows for a greater structural diversity of trees, i.e. greater variation in DBH and height. This may act as a mechanism for niche complementarity, with a canopy of diversely-sized trees able to take advantage of a greater proportion of the available light. \rnew{Additionally, the volume of tree above-ground structures is generally correlated with the volume of below-ground structures \citep{Paul2019}. In water and nutrient limited ecosystems especially, variation in rooting depth may constitute a second related axis of niche partitioning driving the observed positive effect of above-ground structural diversity on AGB \citep{Kulmatiski2013}}. Although we did not measure them here, we would also expect that tree species diversity allows for a greater range of tree functional forms \citep{Pretzsch2014}, i.e. wider variation in canopy shape and overall growth form; broad flat crowns vs. narrow deep crowns, for example. In forests, where the tree canopy is effectively closed, as the stand matures a more diverse canopy emerges via competition and tree mortality events which open canopy gaps \citep{Muscolo2014}. Indeed, our finding that the strength of the effect of tree diversity on AGB increases with stem density supports this mechanism (\autoref{sem_struc_stems_ha}). At low stem densities, competition between mature trees may not occur, meaning that the niche complementarity effect provided by an increase in tree species richness may not be present, accounting for the small effect of tree species diversity on AGB below c. 180 trees ha\textsuperscript{-1}. In frequently disturbed woodlands such as those studied here, a woodland canopy similar to that of a forest is frequently not reached. Instead, a simple open canopy is maintained that can be made more complex and productive via an increase in species diversity.

\rnew{Alternatively, due to the non-linear relationship between biomass and tree size \citep{Bastin2018}, the positive relationship between structural diversity and biomass may also be partly driven by an increased number of large sized trees in plots with higher structural diversity, with large trees contributing disproportionately to biomass. The positive effect of species diversity on AGB via structural diversity may therefore be due to selection effects, with higher diversity plots supporting larger trees due to species specific variation in functional form \citep{Diaz2015}.}

@@

Disturbances such as fire have the potential to reduce both species diversity and above-ground biomass in the short term, due to increased mortality \citep{Huston2014}. Unless this effect is accounted for, there is the potential for mistaken causality as both diversity and biomass may correlate. In our model, time since disturbance is accounted for within each plot via the stem density term. Disturbance reduces stem density of large stems (>10 cm DBH), which is expected to increase until the effects of competition preclude further increase \citep{Johnson2012}. Furthermore, our rarefied measure of species diversity accounts for variation in sampling effort and is therefore independent of stem density. Tree species richness should also increase with time since disturbance as with increased stem density the likelihood of including a new species also increases. Outside of the stem density effect, there are multiple causes for variation in tree species diversity in this study. Vegetation types and localities differ in their available species pool, for example. Variation in abiotic environmental factors will also affect species accumulation.

@@

A negative total effect of soil fertility on AGB is in contrast to other studies in the region and general ecological theory, which predicts a positive effect of soil nutrients on biomass \citep{Scarascia2000}. The negative total effect of soil fertility on AGB was driven mostly by an indirect negative effect via stem density. The direct effect on AGB however, remained positive and marginally significant, as expected. Model estimates of the effect of soil on AGB were poorly constrained compared with other latent variables. This wide standard error on the model predictions is possibly due to the coarseness and nature of the soil data we used. SoilGrids provides modelled data at 250 m resolution, while soil structure and nutrient content varies at much finer scales in southern African woodlands \citep{Muledi2017, Bucini2007}. It is therefore not surprising that this model path is poorly constrained. \citet{Lehmann2014} found similarly weak and poorly constrained relationships for soil in a Structural Equation Model including precipitation, temperature, soil, and fire to predict tree basal area in southern African woodlands. Plot-specific soil data are time-consuming to collect and difficult to compare across studies when different protocols are used. Our study points to the need for further effort in this regard, which may reveal interesting findings about the complex interactions between soil, disturbance and tree diversity in southern African woodlands. Alternatively, \citet{Gourlet2011} found that environmental filtering of fast-growing species with low wood density on resource poor soils resulted in a decoupling of the soil fertility - AGB relationship. It is possible that at regional scales, variation in species composition could offset resource availability constraints on AGB. However, unlike \citet{Gourlet2011} disturbance by fire in our study region may further complicate this environmental filtering effect.

@@

In this study we found that even in highly disturbed southern African woodlands, there exists a generalisable positive association between tree species diversity and ecosystem function, quantified as above-ground woody biomass (AGB). Our findings contribute to our understanding of a universal biodiversity-ecosystem function relationship, one which is moderated in a predictable manner by environmental covariates and their interaction with biodiversity and ecosystem structure. We found that the multiple vegetation types which comprise southern African woodlands exhibit similarities in the relationship between species diversity and woody biomass, suggesting that similar processes operate across the region to determine ecosystem function. We advocate for explicit inclusion of environmental covariates in regional scale models of biodiversity and ecosystem function. We assert that this is necessary to develop our understanding of the biodiversity-ecosystem function relationship in real-world ecosystems, to progress from experimental mesocosms. We found that much of the effect of species diversity on biomass exists as an indirect effect by increasing the structural diversity of trees, exemplifying a key mechanism by which tree species diversity determines ecosystem function in savannas, woodlands and forests, \rchange{i.e. all wooded ecosystems}{where trees comprise a significant, canopy-forming component}. The presence of a stem density threshold above which the effect of tree species diversity on AGB increases clearly implies the presence of niche complementarity effects in southern African woodlands, an aspect which has often been overlooked in previous studies despite its intuitive logic as a determinant of niche complementarity effects in wooded ecosystems. \rchange{Our study shows that biodiversity change through extensive human actions in this region will have the greatest negative impact on ecosystem function in areas of high stem density, and low species diversity, which are those areas predominantly targeted for tree felling.}{Our study shows that biodiversity change through extensive human-induced land use change in this region will have the greatest negative impact on ecosystem function in areas of high stems density, and in certain vegetation types, specifically Mopane and ex-Acacia woodlands.} This raises concerns about the robustness of these ecosystems to further resource extraction and biodiversity loss. Finally, our results provide further evidence of the complex interaction of factors governing biomass and therefore carbon dynamics in disturbance-driven wooded ecosystems, which currently represent the greatest uncertainty in the global terrestrial carbon sink.
